How does character divergence affect resource dynamics?

### 2018-05-20 ECD_Supp_Mat.Rmd

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$
Resource growth is determined by is intrinsic growth rate ($r_i$), carrying capacity ($K_i$), and the feeding rate (or functional response) of each consumer ($F_{ii}(R_i)$, $F_{ij}(R_j)$). Consumer growth is determined by its feeding rate on each resource as well as its conversion efficiency ($e_{ij}$) and mortality rate ($m_i$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size; Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate ($a_{ij}$), such that $F_{ij}(R_j)=a_{ij}R_j$, then our model becomes MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of resource competition.

Here, we examine the effects of this inevitable divergence on resource dynamics.

GIVE DETAILS ABOUT MY ANALYTICAL ANALYSIS AS WELL AS THE TYPE OF SIMULATIONS I PERFORM.

We show that the ecological consequences of character displacement are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

@@

The first foraging scenario we examine is already assumed in MacArthur's model of resource competition, namely that consumers can forage for both resources simultaneously (e.g. they occur in the same habitat). To gain analytical insight to this scenario, we assume that resources are equivalent ($r=r_i$ and $K=K_i$) as well as consumers ($e=e_{ii}=e_{ij}$; $m=m_i$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is simple, it allow us to isolate the effects of character displacement rather than any inherent differences in consumers or resources. In the appendix, we show that when both consumers and resources are present, the abundance of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation:

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource abundances will decrease if consumers evolve higher total attack rates. If there were no constraints on attack rates, each consumer would evolve increasingly higher attack rates on each resource until it goes extinct. In consideration of this, we have included a tradeoff between investments in attacking one resource versus the other such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$; Sargent and Otto 2006). When $n=1$, the tradeoff function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate.

@@

We know from past work that consumers still undergo character divergence regardless of the shape of the tradeoff (Lawlor and Smith 1976; Abrams 1986). However, the shape of the tradeoff qualitatively affects the relationship between character divergence and resource abundance. For example, if consumer's are constrained by a linear tradeoff ($n=1$), such that a unit increase in $a_{ii}$ results in a unit decrease in $a_{ij}$, then character divergence has no effect on resource abundances. If the tradeoff is concave down ($n>1$), then resource abundances can actually increase under character displacement! This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge. Note though that detecting this effect in nature will likely be difficult, since a concave down tradeoff results in relatively small displacement relative to other tradeoff forms (Lawlor and Smith 1976). When the tradeoff is concave up ($n<1$), however, we see that character divergence suppresses resource densities. Although this equilibrium equation for resource abundances was derived for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when there is only one consumer as well (white circle). This is because a consumer, in the absence of competition, evolves to be a generalist that has equal attack rates on each resource, resulting in equivalent resource abundances.

@@

Here we focus on local stability, which makes the most sense in a food-web context. We find that divergence follows the typical "check-mark" stability pattern that has been described recently in theoretical food-web models.

Importantly, this metric of stability allows us to compare the stability of three-species vs. four-species system, which is key for understanding how ecological character displacement affects food-web dynamics.

@@

>>>>

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$

where $r_i$ represents intrinsic growth rates, $K_i$ represents carrying capacity, $e_{ij}$ represents the conversion efficiency of resources into new consumers, and $m_i$ represents the mortality rate of consumers. $F_{ij}(R_j)$ represents the consumer's feeding rate (or functional response) on a specific resource (in this case, $C_i$ on $R_j$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate, such that $F_{ij}(R_j)=a_{ij}R_j$, then our model becomes MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of competition. To be explicit, this means that the specialization of evolved consumers ($\frac{a_{ii}{a_{ii}+a_{ij}}$) is greater in the presence of competitor compared to when no competitor is present.

Here, we examine the effects of this inevitable divergence in attack rates on food-web dynamics. In particular, we examine how divergence in attack rates affects resource abundances as well as the resilience of both resources and consumers to perturbations. We do this by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics and (ii) simulate the effects of competition on the eco-evolutionary dynamics of consumers and resources. For these simulations, we used an Adaptive Dynamics approach. At each evolutionary time step, we created a mutant consumer by randomly choosing a consumer and modifying its attack rate by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. Assuming this mutant consumer was rare, we then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, we updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. We then repeated the simulation 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy (ESS, Price & Maynard Smith 1973) or limit (e.g. $\frac{a_{ii}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). Below is a table of the parameters used in these simulations. Our mathematical analyses show that the consequences of character displacement on food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

## Effect of character divergence on resource abundances

We start by examing the effects of character displacement on the abundance of resources at equilibrium, which will also set the stage for a later analysis of food-web stability. In particular, we examine three different foraging scenarios: (i) consumers forage for resource simultaneously and their feeding rates increase linearly with resource density (MacArthur 1972); (ii) consumers *cannot* forage for resources simultaneously and their feeding rates increase linearly with resource density (Lawlor & Smith 1976); (iii) consumers *cannot* forage for resources simulaneously and their feeding rates on each resource depend on both resource densities and saturate at high densities (McCann 2005).

@@

The first foragine scenario we explore is already assumed in MacArthur's model of resource competition, namely that consumers can forage for both resources simultaneously. In essence, this implies that both resources occur within the same microhabitat. To gain analytical insight to this scenario, we assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is simple, it isolates the process of character displacement, rather than whether these effects depend on inherent differences in consumers or resources. In the appendix, we show that when both consumers and resources are present that the abundance of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation:

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource abundances will decrease if evolution favors an increase in total attack rate. We assume that consumers are not 'Darwinian demons' (Law 1979) and thus the evolution of their attack rates are subject to certain constraints. We modelled these constraints such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$; Sargent and Otto 2006). When $n=1$, the tradeoff function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate.

@@

We know from past work that consumers still undergo character divergence regardless of the shape of the tradeoff (Lawlor and Smith 1976; Abrams 1986). Interestingly, we find that the shape of the tradeoff qualitatively affects the relationship between character divergence and resource abundance. For example, if consumer's are constrained by a linear tradeoff ($n=1$), such that a unit increase in $a_{ii}$ results in a unit decrease in $a_{ij}$, then character divergence has no effect on resource abundances. If the tradeoff is concave down ($n>1$), then resource abundances can actually increase under character displacement! This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge. Note though that detecting this effect in nature will likely be difficult, since a concave down tradeoff results in relatively small displacement relative to other tradeoff forms (Lawlor and Smith 1976). When the tradeoff is concave up ($n<1$), however, we see that character divergence suppresses resource densities. Although this equilibrium equation for resource abundances was derived for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when there is only one consumer (white circle). This is because a consumer, in the absence of competition, evolves to be a generalist that has equal attack rates on each resource, resulting in equivalent resource abundances.

@@

@@

### Other relevant points

Although we used an Adaptive Dynamics approach here, we expect that our results would apply to different models of the genetic architecture of attack rates (e.g. quantitative genetics, Taper & Case 1985), given that character divergence is an inevitable outcome of resource competition, regardless of the model of genetic architecture, when resources dynamics are explicitly modelled (Taper & Case 1985).

These conclusions are only for substitutable resources and that exhibit biological dynamics.
